Consistent Polyhedral Surrogates via Embedding

### 2019-05-22 notes/bo-separated.tex
updated separated proof

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

\begin{lemma} \label{lemma:distance-loss}
  If $L$ is a polyhedral loss, then for each $p$, there exists a constant $c$ such that:
  if the set $U$ is optimal with expected loss $z$, and $d(U,u) \geq \delta$, then $L(u;p) > z + c\delta$.
\end{lemma}
\begin{proof}
  \bo{Changed the statement to something hopefully correct}
  \raf{Correct, but not strong enough for separated link.  Let's chat about it when the rest is settled.}
%  \bo{TODO}
%  \raf{My first thought was ``This is a great lemma!'' but I don't think it's quite true.  Again think of hinge, and take $p\to 0.5$ from above; $U = \{1\}$ the whole time, but $L(-1,p) \to L(1,p)$, so no constant $c$ will work.  We will want a lemma like this, but I think we'll want to relate $\ell$ differences to $L$ differences, especially since that's the new definition of separated link anyway.  (Note that this more nuanced version holds for this hinge example: yes $L(-1,p)$ is getting closer to $L(1,p)$, but so is $\ell(-1,p)$ to $\ell(1,p)$.}
\end{proof}

\begin{theorem}
  Let $L(u,y)$ be a surrogate loss for $\ell(r,y)$ for which there exists some calibrated link.
  Suppose $L$ is a polyhedral loss.
  Then there exists a calibrated, \emph{separated} link $\psi$ between these losses, meaning, for all $p$,
    \[ \inf_{u \in \R' : \psi(u) \not\in \gamma(p)} L(u;p) > \inf_{u \in \R'} L(u;p) . \]
\end{theorem}
\begin{proof}
  We construct $\psi$.

  Fix a small enough $\epsilon$ as promised by Corollary \ref{cor:thick-intersect}.

  We will label each point $u \in \R'$ with a set of legal mappings $r \in \R$.
  Let $\Psi: \R' \toto \R$ be this labeling.
  Begin by initializing $\Psi(u) = \R$ for all $u$, i.e. all mappings are legal.
  Then, we apply the following iterative procedure.
  For each set $U \in \mathcal{U}$, for each $u \in B(U,\epsilon)$, we set $\Psi(u) = \Psi(u) \cap R_U$ where $R_U$ is given by Lemma \ref{lemma:U-surject}.
  Then, construct $\psi$ by making arbitrary legal choices, e.g. $\psi(u)$ is defined as the first element of $\R$ in $\Psi(u)$.

  We have to prove, first, that $\Psi(u)$ is nonempty for each $u$, and second, that $\psi$ is a separated, calibrated link.

  Consider any $u \in \R'$.
  If $u$ is not in $B(U,\epsilon)$ for any $U \in \mathcal{U}$, then we have $\Psi(u) = \R$, so it is nonempty.
  Otherwise, let $\{U_j : j \in \mathcal{J}\}$ be the family whose thickenings intersect at $u$.
  By Corollary \ref{cor:thick-intersect}, because of our choice of $\epsilon$, the family themselves has nonempty intersection.
  By Lemma \ref{lemma:calibrated-pos}, their corresponding report sets $\{R_j : j \in \mathcal{J}\}$ also intersect at some $r$, so $\Psi(u)$ is nonempty.

  Finally, we must show that $\psi$ is a separated, calibrated link.
  It is calibrated because each $U$ maps under $\psi$ into the corresponding $R$ (since its entire thickening does).
  For separation, fix $p$. Now, suppose $U$ is optimal and $u$ is not.
  Now, if $u$ does not map to an optimal report, then it is not in $B(U,\epsilon)$.
  So by Lemma \ref{lemma:distance-loss}, its expected loss is at least $c\epsilon$ larger than optimal for some $c$ depending only on $L$ and $p$.
  This holds for all such $u$, so the link is separated.
\end{proof}

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{corollary}{Corollary}

\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

\begin{definition} \label{def:thick-link}
  Given polyhedral $L$ embedding some $\ell$, the \emph{$\epsilon$-thickened link} $\psi: \R' \to \R \cup \{\bot\}$ is constructed as follows.
  We will label each point $u \in \R'$ with a set of legal mappings $r \in \R$.
  Let $\Psi: \R' \toto \R$ be this labeling.
  Begin by initializing $\Psi(u) = \R$ for all $u$, i.e. all mappings are legal.
  Then, we apply the following iterative procedure.
  For each set $U \in \mathcal{U}$, for each $u \in B(U,\epsilon)$, we set $\Psi(u) = \Psi(u) \cap R_U$ where $R_U$ is given by Lemma \ref{lemma:U-surject}.
  Then, construct $\psi$ by making arbitrary legal choices, e.g. $\psi(u)$ is defined as the first element of $\R$ in $\Psi(u)$.
  If there is no legal choice because $\Psi(u) = \emptyset$, set $\psi(u) = \bot$.
\end{definition}

\begin{theorem} \label{thm:small-eps-thick}
  For all small enough $\epsilon$, the epsilon-thickened link $\psi$ is a well-defined link function from $\R'$ to $\R$, i.e. $\psi(u) \neq \bot$ for all $u$.
\end{theorem}
\begin{proof}
  Fix a small enough $\epsilon$ as promised by Corollary \ref{cor:thick-intersect}.
  Consider any $u \in \R'$.
  If $u$ is not in $B(U,\epsilon)$ for any $U \in \mathcal{U}$, then we have $\Psi(u) = \R$, so it is nonempty.
  Otherwise, let $\{U_j : j \in \mathcal{J}\}$ be the family whose thickenings intersect at $u$.
  By Corollary \ref{cor:thick-intersect}, because of our choice of $\epsilon$, the family themselves has nonempty intersection.
  By Lemma \ref{lemma:calibrated-pos}, their corresponding report sets $\{R_j : j \in \mathcal{J}\}$ also intersect at some $r$, so $\Psi(u)$ is nonempty.
\end{proof}

\begin{lemma} \label{lemma:distance-loss}
  If $L$ is a polyhedral loss, then for each $p$, there exists a constant $c$ such that, for all $u$,
    \[ L(u;p) - \inf_{u^* \in \R'} L(u^*;p) \geq c \cdot d(\Gamma(p),u) . \]
\end{lemma}
\begin{proof}
  Fix $p$ and let $U = \Gamma(p)$.
  If $u \in U$, then both sides are zero.
  So it remains to find a $c$ such that the inequality holds for all $u \not\in U$.

  $L(\cdot;p)$ is a convex polyhedral function, so it is the pointwise maximum over finitely many linear functions.
  Construct the convex polyhedral function $\underbar{L}(\cdot;p)$ by dropping from the max those linear functions that are never equal to $L$ for any $u^* \in U$.
  We have $\underbar{L}(u^*;p) = L(u^*;p)$ for all $u^* \in U$ and $\underbar{L}(u;p) \leq L(u;p)$ for all $u \not\in U$.
  Now $\underbar{L}$ is the max over a finite number of linear functions.
  Each such function $f$ with gradient $\nabla f$ is equal to $\underbar{L}$ above a closed, convex cell in the power diagram formed by projecting $\underbar{L}(\cdot;p)$.
  If $f$ has nonzero gradient, its cell overlaps with $U$ exactly at some face of $U$.
  We will prove that there exists $c_f > 0$ such that, for all $u$ in the corresponding cell of the power diagram,
    \[ \underbar{L}(u;p) \geq L(u^*;p) + c_f \cdot d(U,u) . \]
  We will then repeat this for the finitely many $f$ with nonzero gradient (which covers all points $u \not\in U$) and take the minimum to obtain the desired $c > 0$.

  Consider the set of unit vectors $\{v \in \reals^d : \|v\|=1\}$ and the set of $u^*$ on the boundary of $U$.
  For all pairs $u^*,v$ such that $v$ exposes $u^*$, let
    \[ G_{u^*,v} = \left\{ u^* + \beta v : \beta \geq 0 \right\} . \]
  Note that $\{ u : u \not\in U\} \subseteq \cup_{u^*,v} G_{u^*,v}$.
  We claim each $G_{u^*,v}$ can be associated with some linear $f$ such that $f(u) = \underbar{L}(u;p)$ for all $u \in G_{u^*,v}$.
  \bo{Need to prove this; I think it follows because $G_{u^*,v}$ lies completely within some cell of the power diagram formed by projecting $\underbar{L}$.}
  Hence for all $u$ in $G_{u^*,v}$, we have
  \begin{align*}
    \underbar{L}(u;p) &= \underbar{L}(u^*;p) + (\nabla f) \cdot (d(u^*,u) v)  \\
                      &= L(u^*;p) + c_{u^*,v} \cdot d(U,u)
  \end{align*}
  where $c_{u^*,v} := (\nabla f) \cdot v$, and we note $u^* = \min_{u' \in U} d(u',u)$. \bo{Should prove this; ought to follow because $v$ exposes $u^*$.}
  We must have $c_{u^*,v} > 0$, otherwise $G_{u^*,v} \subseteq U$ which is a contradiction.

  Now for each $f$, we claim the set $\{u^*, v : \text{$G_{u^*,v}$ is associated with $f$}\}$ is closed.
  This follows because the set of $u^*$ such that $f(u^*) = \underbar{L}(u^*;p)$ is closed, and for fixed $u^*$, the set of $v$ such that $G_{u^*,v}$ is in the cell of the power diagram is also closed.\bo{Should prove this.}
  So for this $f$, the infimum over all positive $c_{u^*,v}$ is achieved by some positive member $c_f > 0$.
\end{proof}

\begin{theorem}
  For small enough $\epsilon$, the $\epsilon$-thickened link $\psi$ satisfies that, for all $p$, there exists $\delta > 0$ such that, for all $u \in \R'$,
    \[ L(u;p) - \inf_{u^* \in \R'} L(u^*;p) \geq \delta \left[ \ell(\psi(u);p) - \min_{r^* \in \R} \ell(r^*;p) \right] . \]
\end{theorem}
\begin{proof}
  We take the $\epsilon$ thickened link, which is well-defined by Theorem \ref{thm:small-eps-thick}.
  Fix $p$ and let $U = \Gamma(p)$.
  The left-hand side is nonnegative, so it suffices to prove the result for all $u$ such that the right side is strictly positive, i.e. for all $u$ such that $\psi(u) \not\in \gamma(p)$.
  By definition of the $\epsilon$-thickened link, we must have $d(U,u) \geq \epsilon$.
  By Lemma \ref{lemma:distance-loss}, we have $L(u;p) - \inf_{u^*} L(u^*;p) \geq C$ where $C = c\epsilon$ for some $c > 0$.
  This holds for all $u$.
  Meanwhile,
    \[ \ell(\psi(u);p) - \min_{r^*} \ell(r^*;p) \leq \max_{r \in \R} \ell(r;p) - \min_{r^* \in \R} \ell(r^*;p) =: D, \]
  for some constant $D$.
  This also holds for all $u$.
  Set $\delta = \frac{C}{D}$ to complete the proof.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\break

Bo: this is the proof sketch of the stronger result that Raf and I discussed.

\begin{theorem}
  There exists $\delta$ such that, for all $p,u$,
    \[ L(u;p) - \inf_{u^*} L(u^*;p) \geq \delta \left[ \ell(\psi(u);p) - \min_{r^*} \ell(r^*;p) \right] . \]
\end{theorem}
\begin{proof}[Sketch]
  The above theorem gives a $\delta_p > 0$ for each $p$.
  \bo{Why doesn't this work? $\Delta_{\Y}$ is a closed set, so the infimum delta is positive, done.}

  \bo{The kind of argument we were using...}
  Fix $r^*$ and let $p^*$ be in the relative interior of the level set $\gamma_{r^*}$. (We have conditions implying that $\gamma_{r^*}$ is full-dimensional in the simplex...)
  Let $\{ p^b : b \in B\}$ be the relative boundary of the level set $\gamma_{r^*}$.
  It is indexed by, for example, the closed unit sphere $B = \{w : \|w\|=1\}$ via $p_b = \text{relbound}(\gamma_{r^*}) \cap \{p^* + \beta b : \beta \geq 0\}$.
  For $\alpha \in [0,1]$, define the distribution $p^b_{\alpha} := \alpha p^* + (1-\alpha) p^b$ 
  \begin{claim} \label{claim:p-bound-1}
    Fix $b$.
    There exists $\beta_b \in \reals$ such that $\inf_{u: d(U(p),u) \geq \epsilon} L(u;p^b_{\alpha}) - L(u^*;p^b_{\alpha}) \geq \alpha \beta_b$ for all $\alpha \in [0,1]$.
  \end{claim}
  Given this, we claim that $\beta_{b'} := \inf \{\beta_b : b \in B\}$ is achieved at some $b' \in B$ because it is a closed set.
  So for all $p \in \gamma_{r^*}$, we have $\inf_{u: d(U(p),u) \geq \epsilon} L(u;p) - L(u^*;p) \geq \alpha \beta_{b'}$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bo: notes from chats with Raf about the separated result.

%%%%%%%%% third chat, May 20

%Main Thm: Show: for all r, r': for all p such that r' in gamma(p), we have
%inf_{u : psi(u) = r} L(u;p) - L(u*;p) >= delta * [ ell(r;p) - ell(r';p) ]
%
%Proof: fix p_canon in interior of gamma_{r'}
%Fix p_bd on boundary
%Show exists v such that inf_{u: psi(u) = r} L(u;p) - L(u*;p) >= v*p for all p between p_canon and p_bd
%Implies there exists delta > 0 such that inf_{u: psi(u) = r} L(u;p) - L(u*;p) >= delta * [ell(r;p) - ell(r';p)]
%Argue that p_bd comes from a closed set, so the inf over these delta is achieved at some delta > 0.
%
%
%Weaker lemma: Given r, r', p_canon, p_bd: show exists v such that
%inf_{u: psi(u) = r} L(u;p) - L(u*;p) >= v*p
%p*v > 0 for p in gamma_{r'} \ gamma_r
%p*v = 0 for p in gamma_{r'} \cap gamma_r .
%(for all p between p_canon and p_bd)
%Improved weaker lemma: case on p_bd in gamma_r or not.
%
%%%%%%%%% second chat

%let r1, r2, p be given with r1 in gamma(p).
%
%
%want inf_{u: psi(u) = r2} L(u;p) - L(u*;p) >= delta
%     -------------------------------------
%           ell(r2;p) - ell(r1;p)
%
%for each p, we get a delta > 0.
%
%
%
%
%Fix p. We have an optimal set U and "facets" (cell of the power diagram) {F}.
%
%Let d_f^y = nabla L_y(u_f).
%
%So given p, we have < p, d_f^y > is expected loss (plus something?)??
%
%Expected loss: sum_y p(y) L_y(u_f)
%            >=   sum_y p(y) (c_y + ||d_f^y|| epsilon)
%             =   c_y + epsilon <p, ||d_f||>
%
%L(u_f;p) - L(u_p;p)
%
% sum_y p(y) [ L_y(u_b) + d_f^y * (u_f - u_b) ] - sum_y p(y) L_y(u_b)
%
% = sum_y p(y) [ d_f^y * (u_f - u_b) ]
%>= sum_y p(y) [ delta_y * epsilon ]
%
%
%
%directions of steepest increase (slope)
%
%d_f exposes a face of U ... perhaps the face adjacent to the "facet" f?
%
%U = tall thin triangle example, even with arbitrarily steep slopes, can have arbitrarily low function values, even epsilon away from U!!!

%%%%%%%%%% first chat
%
%for all p, u:
%
%min_{embedded points} < p , L(u) - L(point) >
%
%>= delta * min_{r in R} < p , l(psi(u)) - l(r) >
%
%
%for all p, r' in gamma(p), r' \neq r:
%
%inf_{u : psi(u) = r} <p, L(u) - L(u_{r'})>  >=  delta <p, l(r) - l(r')>
%
%
%L_y(u) - L_y(u_{r'})        L_y(u_r) - L_y(u_{r'})
%
%
%
%Notice that whatever p is, its dot product with l(r) - l(r') is positive.
%So have to show that if p is close to dot-product zero on the left, it's also on the right.
%
%
%
%Draw the line from u_{r'} to u, eventually the slope becomes positive and stays positive for at least epsilon
%
%say d_y = "slope" toward u_{r'}
%
%
%let v = (u-u_{r'}) / ||u-u_{r'}||
%
%d_y = directional derivative in direction v (once positive)
%
%so I have <p, d_y*epsilon >
